The evolution of gene expression in the term placenta of viviparous mammals

### 2015-12-18 manuscript/manuscript.Rnw
load xspace properly
use capitalized crefs
fix ref usage
The pig is actually Duroc, not Wild Boar.
(The testes came from wild boar, and the pig placenta came from a duroc
x duroc cross.)
include bad reads

\documentclass[english,12pt]{article}
\usepackage{fontspec}
\setmainfont{FreeSerif}
\setsansfont{FreeSans}
\setmonofont{FreeMono}
\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,headheight=14pt]{geometry}
\usepackage{url}
\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage[bf]{caption}
\usepackage{rotating}
\usepackage{wrapfig}
\usepackage{booktabs}
\usepackage{float}
\usepackage{multirow}
\usepackage{setspace}
\usepackage{wrapfig}
% \usepackage{txfonts}
%\usepackage[plain]{fancyref}
\usepackage{acronym}
\usepackage{array}
\usepackage{dcolumn}
\usepackage{adjustbox}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage[backend=biber,natbib=true,hyperref=true,style=reading,autocite=inline]{biblatex}
\addbibresource{references.bib}
\usepackage[noblocks,auth-sc]{authblk}
\usepackage[hyperfigures,bookmarks,colorlinks,linkcolor=black,urlcolor=black]{hyperref}
%\usepackage[hyperfigures,bookmarks,colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black]{hyperref}
% Floats at end for journals
% \usepackage[nolists,figuresfirst,nomarkers,heads,nofighead,notabhead]{endfloat}
% \AtBeginFigures{\section*{Figure Legends}\renewcommand{\efloatseparator}{\hbox{}}}
% \AtBeginTables{\section*{Tables}\renewcommand{\efloatseparator}{\mbox{}}}
\usepackage[nomargin,inline,draft]{fixme}
\newcommand{\DLA}[1]{\textcolor{red}{\fxnote{DLA: #1}}}
\usepackage[x11names,svgnames]{xcolor}
\usepackage{texshade}
\usepackage{tikz}
\usepackage{nameref}
\usepackage{zref-xr,zref-user}
\zxrsetup{toltxlabel}
\zexternaldocument*{supplemental_material}
\usepackage{cleveref}
\newenvironment{narrow}[2]{%
  \begin{list}{}{%
      \setlength{\topsep}{0pt}%
      \setlength{\leftmargin}{#1}%
      \setlength{\rightmargin}{#2}%
      \setlength{\listparindent}{\parindent}%
      \setlength{\itemindent}{\parindent}%
      \setlength{\parsep}{\parskip}}%
  \item[]}{\end{list}}
\def\newblock{\hskip}
\newenvironment{paperquote}{%
  \begin{quote}%
     \it
  }%
  {\end{quote}}
\renewcommand{\textfraction}{0.15}
\renewcommand{\topfraction}{0.85}
\renewcommand{\bottomfraction}{0.65}
\renewcommand{\floatpagefraction}{0.60}
%\renewcommand{\baselinestretch}{1.8}
\newenvironment{enumerate*}%
  {\begin{enumerate}%
    \setlength{\itemsep}{0pt}%
    \setlength{\parskip}{0pt}}%
  {\end{enumerate}}
\newenvironment{itemize*}%
  {\begin{itemize}%
    \setlength{\itemsep}{0pt}%
    \setlength{\parskip}{0pt}}%
  {\end{itemize}}
\oddsidemargin 0.0in 
\textwidth 6.5in
\raggedbottom
\clubpenalty = 10000
\widowpenalty = 10000
% make the afill smaller
\renewcommand\Affilfont{\itshape\small}
\renewcommand\cite[1]{\autocite{#1}}
\pagestyle{fancy}

\newcommand{\smonkey}[0]{\textit{A. fusciceps}\xspace}
\newcommand{\cow}[0]{\textit{B. taurus}\xspace}
\newcommand{\dog}[0]{\textit{C. familiaris}\xspace}
\newcommand{\armadillo}[0]{\textit{D. novemcinctus}\xspace}
\newcommand{\horse}[0]{\textit{E. galili}\xspace}
\newcommand{\elephant}[0]{\textit{L. africana}\xspace}
\newcommand{\human}[0]{\textit{H. sapiens}\xspace}
\newcommand{\opossum}[0]{\textit{M. domestica}\xspace}
\newcommand{\mouse}[0]{\textit{M. musculus}\xspace}
\newcommand{\galili}[0]{\textit{N. galili}\xspace}
\newcommand{\sheep}[0]{\textit{O. aries}\xspace}
\newcommand{\bonobo}[0]{\textit{P. paniscus}\xspace}
\newcommand{\carmeli}[0]{\textit{S. carmeli}\xspace}
\newcommand{\boar}[0]{\textit{S. scrofa}\xspace}

@@

<<load.data,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("combined_fpkm","gene_trees","housekeeping_genes_superset","placenta_core_transcriptome","placenta_go_results_bp","placenta_go_results_mf","hugo_nomenclature","gene_annotation")))>>=
for (a in paste0("../data/",c("combined_fpkm","gene_trees",
                              "housekeeping_genes_superset",
                              "placenta_core_transcriptome",
                              "placenta_go_results_bp",
                              "placenta_go_results_mf",
                              "hugo_nomenclature",
                              "gene_annotation"
                              ))) {
    load(a)
}
combined.fpkm[,species:=factor(species,levels=names(species.names))]
setkey(genes.to.tree,"gene_id")
star.logs[,value:=gsub("%$","",value)]
star.logs.wide <- 
    data.table(dcast(star.logs,species+file~field))
setnames(star.logs.wide,
         "Average input read length",
         "input.read.length")
setnames(star.logs.wide,
         "Number of input reads",
         "input.reads")
setnames(star.logs.wide,
         "Uniquely mapped reads number",
         "unique.map.reads")
setkey(hugo.nomenclature,"Approved Symbol")
@

@@

\begin{table}
  \begin{tabular}{c c c c}
    \toprule
    Species                       & Common Name        & Reference & Read Length \\
    \midrule
    \textit{Ateles fusciceps}     & Spider monkey      & \textit{Homo sapiens} & 100 \\
    \textit{Bos taurus}           & Cow                & UMD3.1     & 75 \\
    \textit{Canis familiaris}     & Dog                & CanFam3.1  & 150 \\
    \textit{Dasypus novemcinctus} & 9-Banded Armadillo & DASNOV3    & 100 \\
    \textit{Equus caballus}       & Horse              & EquCab2    & 100 \\ 
    \textit{Loxodonta africana}   & Elephant           & loxAfr3    & 75 \\
    \textit{Homo sapiens}         & Human              & GRCh38     & 75, 150 \\
    \textit{Monodelphis domestica} & Opossum           & BROAD05    & 75 \\
    \textit{Mus musculus}         & Mouse              & GRCm38     & 35,44,150 \\
    \textit{Nannospalax galili}   & Blind mole rat     & \textit{Mus musculus} & 100 \\
    \textit{Ovis aries}           & Sheep              & Oar\_v3.1   & 151 \\
    \textit{Pan paniscus}         & Bonobo             & CHIMP2.1.4 & 150 \\
    \textit{Spalax carmeli}       & Blind mole rat     & \textit{Mus musculus} & 100 \\
    \textit{Sus scrofa}           & Wild Boar          & Sscrofa10.2 & 51 \\
    \bottomrule
  \end{tabular}
  \caption{Reference genomes for aligned species}
  \label{tab:ref_genome_aligned_species}
\end{table}

@@

For transcriptomes sequenced with 75 bp, %
(\Sexpr{species(star.logs.wide[input.read.length > 70 & input.read.length < 80,unique(species)])})
%
,
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,max(as.numeric(input.reads))]}
with at least 96\% of reads \DLA{This is wrong}
mapping to the respective reference genome (\cref{tab:ref_genome_aligned_species}). 
%
For transcriptomes sequenced with 150 bp, %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length > 120 & input.read.length < 160,unique(species)])})
%
,
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  > 120 & input.read.length < 160,max(as.numeric(input.reads))]}
%
with at least 99\% of reads mapping to the
respective reference genome \DLA{This is also wrong}.
%
%
For transcriptomes sequenced with 35 bp, %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length < 40 ,unique(species)])})
%
,
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length < 40 ,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  < 40  ,max(as.numeric(input.reads))]}
%
with at least 99\% of reads mapping to the
respective reference genome \DLA{This is also wrong}.
%
All reads were submitted to the NCBI Sequence Read Archive (SRA)
(Accession Number XXX-YYY).

@@

The top 50 non-housekeeping expressed genes as measured by \ac{FPKM}
in the term placenta of \dog are shown in
\cref{tab:top50_dog_placenta} with a list of all genes in Supplemental
\cref{tab:all_canis_familiaris_placenta}. The gene with the highest
expression %
(\ac{FPKM}=%
\Sexpr{gene.expression("canis familiaris","ENSCAFG00000002115")}) %
is the hormone relaxin (\textit{RLN2}). Other highly expressed genes
among the top 50 (\cref{tab:top50_dog_placenta}) include genes
involved in the cytoskeleton (\textit{TMSB10}, \textit{TMSB4X}),
anticoagulation (\textit{TFPI2}, \textit{PLA2G7}), innate immunity
(\textit{LCN2}, \textit{GM2A}), lipid metabolism (\textit{APOE}),
inhibition of matrix metalloproteinases (\textit{TIMP2},
\textit{TIMP3}), and growth factors (\textit{PGF}, \textit{EGFR},
\textit{HTRA1}). The gene \textit{LGALS1}, which shows evidence of
immunosupression during pregnancy
\cite{Than.ea2008:Emergenceofhormonaland}, is highly expressed %
(\ac{FPKM}=\Sexpr{gene.expression("canis familiaris","LGALS1")}). %
In addition, an ortholog of SSLP1 (ENSCAFG00000010638) a gene
expressed in the placentas of \cow (%
\Sexpr{gene.expression("bos tarus","SSLP1")}, %
\cref{tab:top50_cow_placenta}) and \sheep
(\Sexpr{gene.expression("ovis aries","SSLP1")},
\cref{tab:top50_sheep_placenta}) but is not present in primates
\cite{Ushizawa.ea2009:Characterizationandexpressionanalysis}
(\cref{fig:sslp1_evolution_tree}), is also expressed at high levels
%
(\ac{FPKM}=\Sexpr{gene.expression("canis familiaris","ENSCAFG00000010638")}). %
There is no placental expression of the estrogen synthase gene
(\textit{CYP19A1}) in \dog (~\ref{fig:estrogen_evolution_tree}), 
however expression of \textit{HSD3B1} is relatively high
(\ac{FPKM}=\Sexpr{gene.expression("canis familiaris","HSD3B1")}).

>>>>

\documentclass[english,12pt]{article}
\usepackage{fontspec}
\setmainfont{FreeSerif}
\setsansfont{FreeSans}
\setmonofont{FreeMono}
\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,headheight=14pt]{geometry}
\usepackage{url}
\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage[bf]{caption}
\usepackage{rotating}
\usepackage{wrapfig}
\usepackage{booktabs}
\usepackage{float}
\usepackage{multirow}
\usepackage{setspace}
\usepackage{wrapfig}
% \usepackage{txfonts}
%\usepackage[plain]{fancyref}
\usepackage{acronym}
\usepackage{array}
\usepackage{dcolumn}
\usepackage{adjustbox}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage[backend=biber,natbib=true,hyperref=true,style=reading,autocite=inline]{biblatex}
\addbibresource{references.bib}
\usepackage[noblocks,auth-sc]{authblk}
\usepackage[hyperfigures,bookmarks,colorlinks,linkcolor=black,urlcolor=black]{hyperref}
%\usepackage[hyperfigures,bookmarks,colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black]{hyperref}
% Floats at end for journals
% \usepackage[nolists,figuresfirst,nomarkers,heads,nofighead,notabhead]{endfloat}
% \AtBeginFigures{\section*{Figure Legends}\renewcommand{\efloatseparator}{\hbox{}}}
% \AtBeginTables{\section*{Tables}\renewcommand{\efloatseparator}{\mbox{}}}
\usepackage[nomargin,inline,draft]{fixme}
\newcommand{\DLA}[1]{\textcolor{red}{\fxnote{DLA: #1}}}
\usepackage[x11names,svgnames]{xcolor}
\usepackage{texshade}
\usepackage{tikz}
\usepackage{nameref}
\usepackage{xspace}
\usepackage{zref-xr,zref-user}
\zxrsetup{toltxlabel}
\zexternaldocument*{supplemental_material}
\usepackage[capitalize]{cleveref}
\newenvironment{narrow}[2]{%
  \begin{list}{}{%
      \setlength{\topsep}{0pt}%
      \setlength{\leftmargin}{#1}%
      \setlength{\rightmargin}{#2}%
      \setlength{\listparindent}{\parindent}%
      \setlength{\itemindent}{\parindent}%
      \setlength{\parsep}{\parskip}}%
  \item[]}{\end{list}}
\def\newblock{\hskip}
\newenvironment{paperquote}{%
  \begin{quote}%
     \it
  }%
  {\end{quote}}
\renewcommand{\textfraction}{0.15}
\renewcommand{\topfraction}{0.85}
\renewcommand{\bottomfraction}{0.65}
\renewcommand{\floatpagefraction}{0.60}
%\renewcommand{\baselinestretch}{1.8}
\newenvironment{enumerate*}%
  {\begin{enumerate}%
    \setlength{\itemsep}{0pt}%
    \setlength{\parskip}{0pt}}%
  {\end{enumerate}}
\newenvironment{itemize*}%
  {\begin{itemize}%
    \setlength{\itemsep}{0pt}%
    \setlength{\parskip}{0pt}}%
  {\end{itemize}}
\oddsidemargin 0.0in 
\textwidth 6.5in
\raggedbottom
\clubpenalty = 10000
\widowpenalty = 10000
% make the afill smaller
\renewcommand\Affilfont{\itshape\small}
\renewcommand\cite[1]{\autocite{#1}}
\pagestyle{fancy}

\newcommand{\smonkey}[0]{\textit{A. fusciceps}\xspace}
\newcommand{\cow}[0]{\textit{B. taurus}\xspace}
\newcommand{\dog}[0]{\textit{C. familiaris}\xspace}
\newcommand{\armadillo}[0]{\textit{D. novemcinctus}\xspace}
\newcommand{\horse}[0]{\textit{E. galili}\xspace}
\newcommand{\elephant}[0]{\textit{L. africana}\xspace}
\newcommand{\human}[0]{\textit{H. sapiens}\xspace}
\newcommand{\opossum}[0]{\textit{M. domestica}\xspace}
\newcommand{\mouse}[0]{\textit{M. musculus}\xspace}
\newcommand{\galili}[0]{\textit{N. galili}\xspace}
\newcommand{\sheep}[0]{\textit{O. aries}\xspace}
\newcommand{\bonobo}[0]{\textit{P. paniscus}\xspace}
\newcommand{\carmeli}[0]{\textit{S. carmeli}\xspace}
\newcommand{\pig}[0]{\textit{S. scrofa}\xspace}

@@

<<load.data,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("combined_fpkm","gene_trees","housekeeping_genes_superset","placenta_core_transcriptome","placenta_go_results_bp","placenta_go_results_mf","hugo_nomenclature","gene_annotation")))>>=
for (a in paste0("../data/",c("combined_fpkm","gene_trees",
                              "housekeeping_genes_superset",
                              "placenta_core_transcriptome",
                              "placenta_go_results_bp",
                              "placenta_go_results_mf",
                              "hugo_nomenclature",
                              "gene_annotation"
                              ))) {
    load(a)
}
combined.fpkm[,species:=factor(species,levels=names(species.names))]
setkey(genes.to.tree,"gene_id")
star.logs[,value:=gsub("%$","",value)]
star.logs.wide <- 
    data.table(dcast(star.logs,species+file~field))
setnames(star.logs.wide,
         "Average input read length",
         "input.read.length")
setnames(star.logs.wide,
         "Number of input reads",
         "input.reads")
setnames(star.logs.wide,
         "Uniquely mapped reads number",
         "unique.map.reads")
star.logs.wide[,`% bad reads`:=(as.numeric(`% of reads mapped to too many loci`)+
                            as.numeric(`% of reads unmapped: other`)+
                            as.numeric(`% of reads unmapped: too many mismatches`)+
                            as.numeric(`% of reads unmapped: too short`))]
star.logs.wide[,`% good reads`:=100-`% bad reads`]
setkey(hugo.nomenclature,"Approved Symbol")
@

@@

\begin{table}
  \begin{tabular}{c c c c}
    \toprule
    Species                       & Common Name        & Reference & Read Length \\
    \midrule
    \textit{Ateles fusciceps}     & Spider monkey      & \textit{Homo sapiens} & 100 \\
    \textit{Bos taurus}           & Cow                & UMD3.1     & 75 \\
    \textit{Canis familiaris}     & Dog                & CanFam3.1  & 150 \\
    \textit{Dasypus novemcinctus} & 9-Banded Armadillo & DASNOV3    & 100 \\
    \textit{Equus caballus}       & Horse              & EquCab2    & 100 \\ 
    \textit{Loxodonta africana}   & Elephant           & loxAfr3    & 75 \\
    \textit{Homo sapiens}         & Human              & GRCh38     & 75, 150 \\
    \textit{Monodelphis domestica} & Opossum           & BROAD05    & 75 \\
    \textit{Mus musculus}         & Mouse              & GRCm38     & 35,44,150 \\
    \textit{Nannospalax galili}   & Blind mole rat     & \textit{Mus musculus} & 100 \\
    \textit{Ovis aries}           & Sheep              & Oar\_v3.1   & 151 \\
    \textit{Pan paniscus}         & Bonobo             & CHIMP2.1.4 & 150 \\
    \textit{Spalax carmeli}       & Blind mole rat     & \textit{Mus musculus} & 100 \\
    \textit{Sus scrofa}           & Duroc (pig)        & Sscrofa10.2 & 51 \\
    \bottomrule
  \end{tabular}
  \caption{Reference genomes for aligned species}
  \label{tab:ref_genome_aligned_species}
\end{table}

@@

\Cref{fig:transcriptome_statistics} shows the RNAseq read and
alignment statistics for all species. For transcriptomes sequenced
with 75 bp %
%
(\Sexpr{species(star.logs.wide[input.read.length > 70 & input.read.length < 80,unique(species)])})
%
,
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,max(as.numeric(input.reads))]}
with an average of \Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,mean(`% good reads`)]}\% of reads 
mapping to the respective reference genome (\cref{tab:ref_genome_aligned_species}). 
%
For transcriptomes sequenced with 150 bp, %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length > 120 & input.read.length < 160,unique(species)])})
%
,
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  > 120 & input.read.length < 160,max(as.numeric(input.reads))]}
%
with an average of \Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,mean(`% good reads`)]}\% %
of reads mapping to the
respective reference genome (cref{tab:ref_genome_aligned_species}).
%
%
For transcriptomes sequenced with 35 bp, %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length < 40 ,unique(species)])})
%
,
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length < 40 ,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  < 40  ,max(as.numeric(input.reads))]}
%
with at least 99\% of reads mapping to the
respective reference genome \DLA{This is also wrong}.
%
All reads were submitted to the NCBI Sequence Read Archive (SRA)
(Accession Number XXX-YYY).

@@

The top 50 non-housekeeping expressed genes as measured by \ac{FPKM}
in the term placenta of \dog are shown in
\cref{tab:top50_dog_placenta} with a list of all genes in Supplemental
\cref{tab:all_canis_familiaris_placenta}. The gene with the highest
expression %
(\ac{FPKM}=%
\Sexpr{gene.expression("canis familiaris","ENSCAFG00000002115")}) %
is the hormone relaxin (\textit{RLN2}). Other highly expressed genes
among the top 50 (\cref{tab:top50_dog_placenta}) include genes
involved in the cytoskeleton (\textit{TMSB10}, \textit{TMSB4X}),
anticoagulation (\textit{TFPI2}, \textit{PLA2G7}), innate immunity
(\textit{LCN2}, \textit{GM2A}), lipid metabolism (\textit{APOE}),
inhibition of matrix metalloproteinases (\textit{TIMP2},
\textit{TIMP3}), and growth factors (\textit{PGF}, \textit{EGFR},
\textit{HTRA1}). The gene \textit{LGALS1}, which shows evidence of
immunosupression during pregnancy
\cite{Than.ea2008:Emergenceofhormonaland}, is highly expressed %
(\ac{FPKM}=\Sexpr{gene.expression("canis familiaris","LGALS1")}). %
In addition, an ortholog of SSLP1 (ENSCAFG00000010638) a gene
expressed in the placentas of \cow (%
\Sexpr{gene.expression("bos tarus","SSLP1")}, %
\cref{tab:top50_cow_placenta}) and \sheep
(\Sexpr{gene.expression("ovis aries","SSLP1")},
\cref{tab:top50_sheep_placenta}) but is not present in primates
\cite{Ushizawa.ea2009:Characterizationandexpressionanalysis}
(\cref{fig:sslp1_evolution_tree}), is also expressed at high levels
%
(\ac{FPKM}=\Sexpr{gene.expression("canis familiaris","ENSCAFG00000010638")}). %
There is no placental expression of the estrogen synthase gene
(\textit{CYP19A1}) in \dog (\cref{fig:estrogen_evolution_tree}), 
however expression of \textit{HSD3B1} is relatively high
(\ac{FPKM}=\Sexpr{gene.expression("canis familiaris","HSD3B1")}).
